Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-12-05 doc/manuscript/translocation.qmd
Edits primarily in response to Reviewer 2 comments

---
bibliography: translocation.bib
csl: pnas.csl
# format: docx
format :
  # latex:
  pdf:
#    cite-method: natbib
    include-in-header:
      text: |
        \usepackage{blkarray}
editor: visual
editor_options:
  markdown:
    wrap: sentence
---

@@

One consequence of human alteration of the biosphere is an increase in emerging infectious diseases [@jones2008; @fisher2012].
Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa, including echinoderms, mammals, birds, and amphibians [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
Amid this global loss of biodiversity, some populations decimated by disease have subsequently recovered to varying extents [@scheele2017; @voyles2018; @knapp2016].
Although we know relatively little about the mechanisms underlying recovery [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions].
However, linking these evolutionary responses to population recoveries is difficult and few clear examples exist.
In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [@mendelson2019; @brannelly2021].
To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.

@@

Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant/tolerant genotypes [@knapp2016].
Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].

In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and a detailed population model to provide an unprecedented example of how evolution in response to disease allows the successful reestablishment of extirpated MYL frog populations.
Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide a genetic basis for the increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
We then use the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys to describe the dynamics and drivers of population establishment.
These data are also used to parameterize a population model that extends our inferences of population establishment well beyond the temporal extent of the CMR data.
Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant/tolerant animals can allow the large-scale recovery of disease-impacted species.

@@

Estimates of frog survival during the 1-year period following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens).
This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate), which we suggest resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R. sierrae*.
Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter) [@joseph2018; @vredenburg2010].

@@

To determine whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection, we developed linear and multilevel models.
The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens).
Important predictors included winter severity in the year following translocation, site elevation, and donor population (@fig-cond-effects).
Males had somewhat higher survival than females, but this effect was only marginally important (@fig-cond-effects, @fig-survival-postdens).
Collectively, these results are consistent with frogs in recovering populations having evolved sufficient resistance/tolerance to minimize the adverse effects of Bd infection.
In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs).
Therefore, frog resistance/tolerance in non-natal habitats is maintained across generations.

@@

Results from frog translocations indicated that adult survival was often relatively high and that most populations showed evidence of successful reproduction and recruitment (described above).
Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
To extend our inferences of population establishment beyond those possible from the 2-16 years of site-specific CMR data, we developed a population viability model.
Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values).

Given observed yearly adult survival probabilities (from analysis of CMR data) and a yearly survival probability of the year 1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
As year 1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of eight of twelve population was greater than 1 (@fig-viability A).

Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year 1 juvenile survival and recruitment (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year 1 juvenile survival was greater than 0.10 (@fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in survival and recruitment.
In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \>50% regardless of the value of mean year 1 juvenile survival between 0 - 0.25.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).

@@

@@

Reintroduction of resistant/tolerant MYL frogs was remarkably successful in reestablishing viable populations in the presence of Bd.
Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
Bd loads measured before and after translocation were similar and generally remained well below the level at which frog mortality occurs.
Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined).
The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs.
In addition, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance/tolerance has a genetic basis.
Results from the population viability model were also encouraging.
In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction $<$ 0.5 when year 1 juvenile survival probability was greater than 0.10).
In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant/tolerant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.

Despite the resistance/tolerance of adult MYL frogs in recovering populations, individual and population-level impacts of Bd are still evident.
In an earlier study of 2 of our 12 translocated populations [@joseph2018], Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in @joseph2018 are identified as "70550" and "70505" in the current study).
Applying similar analyses to all 12 of our translocated populations would likely provide a broader perspective of the ongoing effect of Bd.
In addition to these important but relatively subtle effects of Bd on adults, the impacts on younger life stages are more apparent.
MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2018] and as a result experience high mortality.
This high susceptibility of metamorphs is documented in numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
In naturally recovering and translocated MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
Therefore, although adult MYL frogs appear relatively resistant/tolerant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].

@@

@@

For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
The populations that we selected contained several hundred *R. sierrae* adults and hundreds or thousands of tadpoles.
These relatively high abundances were the result of recent increases following previous Bd-caused declines [@knapp2016].
As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 copies)[@vredenburg2010; @joseph2018].
Recipient sites to which frogs were translocated were chosen based on previous *R. sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species [@knapp2005].

We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap, @fig-translocation-survival) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
Frogs were transported to the recipient site either on foot or via helicopter.
Following release, translocated populations were visited approximately once per month during the summer active season (median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
CMR surveys allowed estimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (median number of secondary periods per primary period = `r round(robust_year_median)`; range = `r robust_year_min`--`r robust_year_max`), and frogs were processed as described above.
However, when the number of frogs detected on the first survey day was zero or near zero, we typically conducted only a single-day CMR survey.
When using a robust design, within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.

For each of the translocated populations, we used the presence of tadpoles and/or juveniles from VES and counts of new recruits (i.e, untagged adults) in CMR surveys to provide two measures of successful reproduction.
To calculate the proportion of years in which tadpoles/juveniles were present at a site, we excluded surveys conducted in the year of the initial translocation to that site.
This exclusion accounted for the fact that all translocations were conducted after the breeding period and reproduction would therefore not have occurred until the following year.
Similarly, to calculcate the proportion of years in which new recruits were present at a site, we excluded surveys conducted during the 3 years following the initial translocation.
This accounted for the multi-year tadpole and juvenile stages in MYL frogs ([@tbl-param_values]).

We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
The possible states of individuals include "not recruited", "alive", and "dead".
The possible observations of individuals include "detected" and "not detected".
We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).

>>>>

---
bibliography: translocation.bib
csl: pnas.csl
format: pdf
# format: docx
# format :
#  latex:
#    cite-method: natbib
editor: visual
editor_options:
  markdown:
    wrap: sentence
---

<!--# Drop "evolutionary rescue" from title? -->

@@

One consequence of human alteration of the biosphere is an increase in emerging infectious diseases [@jones2008; @fisher2012].
Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa, including echinoderms, mammals, birds, and amphibians [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
Amid this global loss of biodiversity, some populations decimated by disease have subsequently recovered to varying extents [@scheele2017; @voyles2018; @knapp2016].
Although we know relatively little about the mechanisms underlying recovery [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions]. However, linking these evolutionary responses to population recoveries is difficult and few clear examples exist. 
In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [@mendelson2019; @brannelly2021].
To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed. 
(In this paper, we do not distinguish between resistance and tolerance because although these terms have different meanings and consequences for host-pathogen dynamics, these distinctions have little significance for our primary focus, i.e., describing recovery of MYL frogs in the presence of Bd.) <!--# Feels clumsy to include this parenthetically, but I have not yet come up with a better alternative.  -->

@@

Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
Frogs in these recovering populations show reduced susceptibility to Bd infection [@knapp2016], with infection intensities ("loads") of adults consistently in the low-to-moderate range regardless of frog size [@ellison2019], frog density [@joseph2018; @wilber2022], or time of year [@knapp2011].
This reduced susceptibility is evident even under controlled laboratory conditions [@knapp2016], and is consistent with several non-mutually exclusive mechanisms.
These include natural selection for more resistant/tolerant genotypes [@savage2016], acquired immunity [@grogan2018], and inherent between-populations differences in frog susceptibility that pre-date Bd exposure. 
Of these possibilities, evolution of resistance/tolerance would have the most general relevance for amphibian recovery from the effects of Bd, and as such additional investigation of this potential mechanism is critically needed.
Exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated following Bd epizootics provide a rare, but preliminary, indication that population establishment is possible despite ongoing Bd infection [@joseph2018].

<!--# I've reworded the next paragraph to provide explicit objectives of each of the sections (plus summary of results). -->

In the current study, we had three primary objectives.
First, to determine whether the reintroduction of resistant/tolerant MYL frogs obtained from populations recovering from Bd-caused declines [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
Second, to extend our inferences of population recovery well beyond the temporal extent of our multi-year survey data, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period.
Third, we used exome capture methods to provide insights into the evolutionary response by MYL frogs to Bd infection, and thereby provide a mechanism by which frogs reintroduced into sites from which they were previously extirpated by Bd could recover (instead of being driving to extirpation again).
Our results indicate the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, that the recovering populations are likely to persist over a 50-year period, and that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.

<!--# Previous final paragraph of Introduction (revised post-review to reflect reordered sections): In the current study, we combine results from a 15-year landscape-scale frog reintroduction effort, a detailed population model, and genomic analyses to provide an unprecedented example of how reintroduction of resistant/tolerant frogs obtained from recovering populations [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease. Specifically, the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys indicate the capacity of reintroduced populations to become established and eventually recover. We used these data to parameterize a population model that extends our inferences well beyond the temporal extent of the CMR data, and shows that the recovering populations are likely to persist over a 50-year period. Finally, genomic results obtained using exome capture methods highlight evolutionary rescue as potentially key mechanism by which MYL frogs are able to recover in the presence of ongoing Bd infection. These results show that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease. Collectively, these results provide one of the first examples of how evolutionary rescue might facilitate the successful reintroduction of resistant/tolerant animals and allow the large-scale recovery of disease-impacted species.-->

@@

<!--# I moved this paragraph to the top of results to provide an early focus on outcomes that are more obviously related to population recovery than survival. -->

@@

Estimates of frog survival during the 1-year period following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens).
This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate), which we suggest resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R. sierrae* (see **Materials and Methods - Frog population recovery - Field methods** for details).
Finally, Bd loads were fairly consistent before versus after translocation and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (@fig-bdload-beforeafter) [@joseph2018; @vredenburg2010].

To determine whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection, we developed linear and multilevel models.
The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens).
Important predictors included winter severity in the year following translocation, site elevation, and donor population (@fig-cond-effects).
Males had somewhat higher survival than females, but this effect was only marginally important (@fig-cond-effects, @fig-survival-postdens).
Collectively, these results are consistent with frogs in recovering populations having evolved sufficient resistance/tolerance to minimize Bd load and/or the adverse effects of Bd infection.
In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs).
Therefore, frog resistance/tolerance in non-natal habitats is maintained across generations.

@@

Results from frog translocations indicated that most populations showed evidence of successful reproduction and recruitment and that adult survival was often relatively high (described above).
Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
To extend our inferences of population establishment beyond those possible from the 2-16 years of site-specific CMR data, we developed a population viability model.
Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values).

Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability $\omega$ of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).

Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of 0 to 0.32 (@fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment.
When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50-year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).

@@

To determine whether MYL frog populations show an evolutionary response to Bd, we compared frog genomes between populations with contrasting histories of Bd exposure.
Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
Bd-exposure histories of the nine study populations are based on 10-20 years of visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions [during which Bd loads remain very low; wilber2022], and no history of Bd epizootics since we first first surveyed these populations in the late 1990s and early 2000s [@zhou2015]. 
In contrast, recovering populations exist in an enzootic state [@briggs2010], characterized by relatively small numbers of adults (generally < 100), high Bd prevalence (often > 80%; [@knapp2011], and, in adults, moderate Bd loads that are typically well below the level expected to cause mortality [@vredenburg2010]. 
Naive and recovering populations can be identified unambiguously using these differences in Bd prevalence and load. 
(See SI Appendix for additional details regarding the study design.)  
<!--In response to Editor - Comment 3, mention that there is no dispersal between extant populations due to presence of fish and fish-induced habitat fragmentation?-->

@@

Reintroduction of resistant/tolerant MYL frogs was remarkably successful in reestablishing viable populations in the presence of Bd.
Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
Bd loads measured before and after translocation were similar and generally remained well below the level at which frog mortality occurs.
Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined).
The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs <!--# Add to this to deal with Reviewer 2-Comment 4.-->.
In addition, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance/tolerance has a genetic basis.
Results from the population viability model were also encouraging.
In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction = 0 - 0.32).
In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant/tolerant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
<!--# Mention that 1-year survival can be used as an accurate proxy of long-term persistence.  -->

Despite the resistance/tolerance of adult MYL frogs in recovering populations, individual and population-level impacts of Bd are still evident.
In an earlier study of 2 of our 12 translocated populations [@joseph2018], Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in @joseph2018 are identified as "70550" and "70505" in the current study).
Applying similar analyses to all 12 of our translocated populations would likely provide a broader perspective of the ongoing effect of Bd.
In addition to these important but relatively subtle effects of Bd on adults, the impacts on younger life stages are more apparent.
MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2019] and as a result experience high mortality.
This high susceptibility of metamorphs is documented in numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
In naturally recovering and translocated MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
Therefore, although adult MYL frogs appear relatively resistant/tolerant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].

@@

For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
The populations that we selected contained several hundred *R. sierrae* adults and hundreds or thousands of tadpoles.
These relatively high abundances were the result of recent increases following previous Bd-caused declines [@knapp2016].
As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 copies)[@vredenburg2010; @joseph2018].
Recipient sites to which frogs were translocated were chosen based on previous *R. sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species [@knapp2005].
At the beginning of this study, we had a relatively limited understanding of these habitat quality characteristics. 
In subsequent years, we improved our site selection process by incorporating new information about important habitat features (in particular, overwinter habitats such as submerged boulders and overhanging banks). *R. sierrae* were absent from all recipient sites prior to the first translocation.

We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap, @fig-translocation-survival) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
Frogs were transported to the recipient site either on foot or via helicopter.
Following release, translocated populations were visited approximately once per month during the summer active season (generally July-August; median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
CMR surveys allowed estimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (median number of secondary periods per primary period = `r round(robust_year_median)`; range = `r robust_year_min`--`r robust_year_max`), and frogs were processed as described above.
However, when the number of frogs detected on the first survey day was zero or near zero, we typically conducted only a single-day CMR survey.
When using a robust design, within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.

For each of the translocated populations, we used the presence of tadpoles and/or juveniles from VES and counts of new recruits (i.e, untagged adults) in CMR surveys to provide two measures of successful reproduction.
To calculate the proportion of years in which tadpoles/juveniles were present at a site, we excluded surveys conducted in the year of the initial translocation to that site.
This exclusion accounted for the fact that all translocations were conducted after the breeding period and reproduction would therefore not have occurred until the following year.
Similarly, to calculate the proportion of years in which new recruits were present at a site, we excluded surveys conducted during the 3 years following the initial translocation.
This accounted for the multi-year tadpole and juvenile stages in MYL frogs ([@tbl-param_values]).

We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
The possible states of individuals include "not recruited", "alive", and "dead".
The possible observations of individuals include "detected" and "not detected".
We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
We also assume that new recruits were a result of within-site reproduction and not immigration from adjacent populations because no *R. sierrae* populations were present within several kilometers of the translocation sites. 
For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).

$$
\begin{bmatrix}
L_1 \\
L_2 \\
L_3 \\
J_1 \\
J_2 \\
A_R \\ 
A_T
\end{bmatrix}(t + 1) = 
\begin{bmatrix}
  0 & 0 & 0 & 0 & 0 & \sigma_{A_R} p_F F & \sigma_{A_T} p_F F \\
  \sigma_{L1} p_{L1} & 0 & 0 & 0 & 0 & 0 & 0 \\
  0 & \sigma_{L2} p_{L2} & 0 & 0 & 0 & 0 & 0\\
  \sigma_{L1} (1 - p_{L1}) & \sigma_{L2} (1 - p_{L2}) & \sigma_{L3} & 0 & 0 & 0 & 0 \\
  0 & 0 & 0 & \sigma_{J1} p_{J1} & 0 & 0 & 0 \\
  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) \omega & \sigma_{J2} \omega & \sigma_{A_R} & 0 \\
  0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
\end{bmatrix} \begin{bmatrix}
L_1 \\
L_2 \\
L_3 \\
J_1 \\
J_2 \\
A_R \\
A_T \\
\end{bmatrix}(t) 
$$

The parameters in this model are yearly survival probability $\sigma_{\cdot}$ (the subscript "$\cdot$" indicates a particular state variable), probability that a female frog reproduces in a given year $p_F$, number of eggs produced by a female frog in a year that successfully hatch $F$, probability of a first-year tadpole remaining as a tadpole $p_{L1}$, probability of a second-year tadpole remaining as a tadpole $p_{L2}$, and probability of a first-year juvenile remaining as a juvenile $p_{J1}$, and the probability of a juvenile successfully recruiting to an adult $\omega$.

After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values]), we performed four analyses.
First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
Second, we compute the elasticity of $\lambda$ to model parameters, to quantify how much changes in adult survival, for example, affected the long-run growth rate.
Third, we included demographic and environmental stochasticity in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
Additional details are provided in the SI Appendix.
Code to replicate the analyses can be found at <https://github.com/SNARL1/translocation>.

@@
